### 2012-09-02 sections/conclusion.tex
* implementation   * nfql architecture (with figure)   * flow query intermediate format     * citation to json-c     * flow query and branch structs (listing)     * efficient rule processing using dedicated comparators (listing)   * splitter   * filter     * nfql vs nfql (optimized) (figure)   * grouper     * nfql (generic) vs nfql (specific) (figure)     * group aggregations merged in this sub section   * group filter, merger, ungrouper   * performance evaluations

We presented \ac{NFQL}, an efficient C implementation of the stream-based flow
query language. The language allows applying absolute (or relative) filters,
aggregating flows into groups, evaluating intricate timing relationships among
them, and merging them into one collection. \ac{NFQL} can execute such complex
query tasks in matter of minutes, thereby expanding the scope of current flow
record processing tools. The conducted performance evaluations reveal that
\ac{NFQL} is on par with tools that support only absolute filters. SiLK, the
only package that provides tools that are similar to the rest of \ac{NFQL}'s
processing pipeline appears faster. This is because it can optimize its
operations in favor of the limited set of comparisons that are only based on
equality, and its usage of a proprietory file storage format.
\todo{...}

>>>>

We presented \ac{NFQL}, an efficient C implementation of the stream-based flow
query language. The language allows applying absolute (or relative) filters,
aggregating flows into groups, evaluating intricate timing relationships among
them, and merging them into one collection. \ac{NFQL} can execute such complex
query tasks in matter of minutes, thereby expanding the scope of current flow
record processing tools. The conducted performance evaluations reveal that
\ac{NFQL} is on par with tools that support only absolute filters. SiLK, the
only openly available package that provides tools that are similar to the rest
of \ac{NFQL}'s processing pipeline appears faster. This is because it can
optimize its operations in favor of the limited set of comparisons that are
only based on equality, and its usage of a different file storage format.
\todo{...}

### 2012-09-02 sections/evaluation.tex
* implementation   * nfql architecture (with figure)   * flow query intermediate format     * citation to json-c     * flow query and branch structs (listing)     * efficient rule processing using dedicated comparators (listing)   * splitter   * filter     * nfql vs nfql (optimized) (figure)   * grouper     * nfql (generic) vs nfql (specific) (figure)     * group aggregations merged in this sub section   * group filter, merger, ungrouper   * performance evaluations

We used a public flow trace containing $20M$ records to run a number of flow
queries using our benchmarking suite. We used the publically available Trace 7
from SimpleWeb \cite{simpleweb} repository for all the performance
evaluations. The input trace was compressed at \texttt{ZLIB\_LEVEL} $5$ using
the \texttt{zlib} suite. It was also converted to \texttt{nfdump} and SiLK
compatible formats and compressed with \texttt{zlib} keeping the same
compression level. The suite was run on a high-end
machine\footnote{\url{crystal.eecs.jacobs-university.de}} with $24$ cores of
$2.5$ GHz clock speed and $18$ GiB of memory.

The first set of queries attempt to stress the filter stage.  We use varying
values on the packet field offset to determine the amount of flow records that
are passed by the filter. The resultant filtered records are written to
\texttt{flow-tools} compatible file format and compressed at
\texttt{ZLIB\_LEVEL} $5$ using \texttt{zlib} suite. The ratio of the number of
filtered records in the output trace to the number of the flow records in the
input trace is plotted against time. The evaluation results are shown in
Fig. \ref{fig:benchmarks-filter}.

@@

It can be seen that the performance of the filter stage in \ac{NFQL} is
comparable to that of \texttt{flowtools} and SiLK. SiLK takes less time on
lower ratios, but then again SilK and \texttt{nfdump} also use their own
proprietary format for trace files. As a result, the amount of data that needs
to read (or written) may be different to what it is for \ac{NFQL} and
\texttt{flowtools}. On the other hand, \texttt{nfdump} appears to be
significantly faster than the rest. This is because \texttt{nfdump} lacks
\texttt{zlib} support, and as such the files that are read and written used
\texttt{lzo} compression scheme which trades space for achieving faster
compression and decompression. It is important to note, that all the tools
were single-threaded in this evaluation, and did not completely utilize the
$24$ cores that were at their disposal. It comes as a realization, that
filtering the input using multiple threads by memory mapping the trace and
adding \texttt{lzo} compression will drastically improve \ac{NFQL}'s filter
performance.

The second set of queries attempt to stress the grouper stage. We reuse the
filter query that produces a $1.0$ ratio to allow the grouper to receive the
entire trace as a filtered recordset. The grouper part of the query then
gradually increases the number of grouping terms in the \ac{DNF} expression to
increase the output/input ratio. The resultant groups are again written as
\texttt{flowtools} files using the same \texttt{zlib} compression level. The
ratio of the number of groups formed to the number of the input filtered
records is plotted against time. \texttt{nfdump} and \texttt{flowtools} do not
support grouping, and therefore are not considered in this evaluation. The
evaluation results are shown in Fig. \ref{fig:benchmarks-grouper}.

@@

The evaluation graph reveals that the performance of the \ac{NFQL} grouper
stage is close. The time taken by the tools are comparable on lower ratios,
but on higher ratios, \ac{NFQL} starts to drift apart. Since most of the time
is taken in writing the records to files, it is unclear whether SiLK's usage
of a proprietary format which may reduce reads/writes is responsible for the
drift on higher ratios. SiLK's \texttt{rwgroup} tool is also supplied a
\texttt{--summarize} flag in all the queries. This gives SiLK the leverage to
not store information about which members are part of the group.  \ac{NFQL} on
the other hand needs to allocate resources (which may take time) to keep this
information in its data structures, since the ungrouper later may request to
write the members of a group while unfolding the tuples. It is also important
to note that both the tools again remained single-threaded throughout the
evaluation. SiLK took an advantage of an inherent concurency arising from how
the query is structured as one single \texttt{bash} script using pipes. The
pipe between \texttt{rwsort} and \texttt{rwgroup} makes the two process run
concurrently, the effect of which gets more pronounced on higher ratios and
can be a drift determining factor.  The profiling results from GNU
\texttt{gprof} \cite{graham:1982} indicate that $60\%$ of the time is taken in
\texttt{qsort} comparator calls.  As a result, it comes as no surprise, that
bifurcating \texttt{qsort} invocation to multiple threads and later merging
the results back using merge sort will help parallelize the grouper stage and
maybe reduce the drift on higher ratios. In addition, since all of the
evaluation queries had grouping terms using an equality comparator, \ac{NFQL}
can introspect such a grouping rule to dynamically optimize processing
searches using a hashtable and turn to \texttt{qsort} based grouping only as a
fallback.

The third set of queries attempt to stress the group filter stage. We reuse
the filter and grouper queries that produce a $1.0$ ratio to allow the group
filter to receive the entire trace as input. This means, each flow record of
the original trace now becomes a group record for the group filter.  The group
filter then reuses the same varying values of the packet offset to determine
the amount of groups that are filtered ahead. The filtered groups are again
written as \texttt{flowtools} files using the same \texttt{zlib} compression
level. The ratio of the number of filtered groups formed to the number of the
input group records is plotted against time. The evaluation results are shown
in Fig. \ref{fig:benchmarks-groupfilter}.

@@

It can be seen that the timings of \ac{NFQL} are far apart from that of SiLK.
It is due to the drift already created by the grouper at $1.0$ ratio in the
previous stage. As a result, the group filter comes into play only after $300$
seconds, whereas SiLK's group filtering already starts just below $150$
seconds. Even if we normalize the graph, it can be observed that the group
filter has a significantly higher slope. This is because it is only executed
once the grouper returns, and therefore has to reiterate the groups to make a
filtering decision.

@@

It can be seen that the merger is the most performance hit and time consuming
stage of the \ac{NFQL} pipeline thus far. It is due to the fact that the
merger is working on twice the number of flow records than any other previous
stage. In addition, each branch is writing the results of the filter, grouper
and group filter stage to \texttt{flowtools} files. As a result, the amount of
disk I/O involved is twice as much as well. Even though each branch is
delegated to a separate core using affinity masks, most of time is taken in
writing these results to the file. These results although look less promising,
they are way better than the previous merger implementation. The newer merger
takes advantage of sorted nature of filtered groups and therefore can
signifantly reduce the number of merger matches. It also writes a merged group
record to file only once despite the number of times it has matched. Without
these optimizations, running such queries on the merger would keep the CPU
churning for days without results.

The last set of queries attempt to stress the ungrouper stage. They reuse the
entire merger queries as is, but enable the ungrouper now as well. This means,
that the ungrouper now attempts to unfold the merged groups returned by the
merger to write their member flow records to \texttt{flowtools} files.
However, since the merger receives each flow record as its own filtered group,
each merged group has only one member. As a result, the ungrouper ends up
executing its logic and rewriting the merged groups as \texttt{flowtools}
files using the using the same \texttt{zlib} compression level. The ratio of
the number of result flow records to twice the number of the flow records in
the input trace is plotted against time. The evaluation results are shown in
Fig. \ref{fig:benchmarks-ungrouper}.

@@

It can be seen that the evaluation behavior is similar to that shown by the
merger, but is just more involved. This is because the ungrouper also has to
write the entire merged groupset to file, as is done by merger. This is the
reason for the execution engine in taking twice the amount of time. In fact,
this evaluation is performed only to stress the functioning of the ungrouper
and running queries that merge each filtered group filtered individually is
less useful in practise. Ideally, the execution engine must shutdown the
ungrouper if the merger merges every filtered group record in its output. This
is because, such a behavior implicitly eliminates the need of the ungrouper.

>>>>

We used a public flow trace containing $20M$ records to run a number of flow
queries using our benchmarking suite. We used a publically available Trace 7
from the SimpleWeb \cite{simpleweb} repository for all the performance
evaluations. The input trace was compressed at \texttt{ZLIB\_LEVEL} $5$ using
the \texttt{zlib} suite. It was also converted to \texttt{nfdump} and SiLK
compatible formats and compressed with \texttt{zlib} keeping the same
compression level. The suite was run on a high-end machine with $24$ cores of
$2.5$ GHz clock speed and $18$ GiB of memory.

The first set of queries attempt to stress the filter stage.  We use varying
values on the \texttt{packet} field offset to determine the amount of flow
records that are passed by the filter. The resultant filtered records are
written to \texttt{flow-tools} compatible file format and compressed at
\texttt{ZLIB\_LEVEL} $5$. The ratio of the number of filtered records in the
output trace to the number of the flow records in the input trace is plotted
against time. The evaluation results are shown in Fig.
\ref{fig:benchmarks-filter}.

@@

It can be seen that the performance of the filter stage in \ac{NFQL} is
comparable to that of \texttt{flowtools} and SiLK. SiLK takes less time on
lower ratios, but then again SilK and \texttt{nfdump} also use their own file
format. As a result, the amount of data that needs to read (or written) may be
different to what it is for \ac{NFQL} and \texttt{flowtools}.  On the other
hand, \texttt{nfdump} appears to be significantly faster than the rest. This
is because \texttt{nfdump} lacks \texttt{zlib} support, and as such the files
that are read and written used \texttt{lzo} compression scheme which trades
space for achieving faster compression and decompression. It is important to
note, that all the tools were single-threaded in this evaluation, and did not
completely utilize the $24$ available cores. It comes as a realization, that
adding \texttt{lzo} compression will drastically improve \ac{NFQL}'s filter
performance.

The second set of queries attempt to stress the grouper stage. We reuse the
filter query that produces a $1.0$ ratio to allow the grouper to receive the
entire trace as a filtered recordset. The grouper part of the query then
gradually increases the number of grouping terms in the \ac{DNF} expression to
increase the output/input ratio. The resultant groups are again written as
\texttt{flowtools} files using the same \texttt{zlib} compression level. The
ratio of the number of groups formed to the number of the input filtered
records is plotted against time. \texttt{nfdump} and \texttt{flowtools} do not
support grouping, and therefore are not considered in this evaluation. The
results are shown in Fig. \ref{fig:benchmarks-grouper}.

@@

The evaluation graph reveals that the performance of the \ac{NFQL} grouper
stage is close to the performance of SiLK. The time taken by the tools are
comparable on lower ratios, but on higher ratios, \ac{NFQL} starts to drift
apart. Since most of the time is taken in writing the records to files, it is
unclear whether SiLK's usage of a proprietary format, which may reduce
reads/writes is responsible for the drift on higher ratios. SiLK's
\texttt{rwgroup} tool is also supplied a \texttt{--summarize} flag to force it
to write only the first record of each group, to make both tools write the
same number of records to files. This gives SiLK the leverage to not store
information about which members are part of the group. \ac{NFQL} on the other
hand needs to allocate resources (which may take time) to keep this
information in its data structures, since the ungrouper later may request to
write the members of a group while unfolding the tuples. It is also important
to note that both the tools again remained single-threaded throughout the
evaluation. SiLK took an advantage of an inherent concurency arising from how
the query is structured as one single \texttt{bash} script using pipes. The
pipe between \texttt{rwsort} and \texttt{rwgroup} makes the two processes run
concurrently, the effect of which gets more pronounced on higher ratios and
can be a drift determining factor.  The profiling results from GNU
\texttt{gprof} \cite{graham:1982} indicate that $60\%$ of the time is taken in
\texttt{qsort} comparator calls.  As a result, it comes as no surprise, that
bifurcating \texttt{qsort} invocation to multiple threads and later merging
the results back using merge sort will help parallelize the grouper stage and
maybe reduce the drift on higher ratios. In addition, since all of the
evaluation queries had grouping terms using an equality comparator, \ac{NFQL}
can introspect such a grouping rule to dynamically optimize processing
searches using a hashtable and turn to \texttt{qsort} based grouping only as a
fallback.

The third set of queries attempt to stress the group filter stage. We reuse
the filter and grouper queries that produce a $1.0$ ratio to allow the group
filter to receive the entire trace as input. This means, each flow record of
the original trace now becomes a group record for the group filter.  The group
filter then reuses the same varying values of the \texttt{packet} field offset
to determine the amount of groups that are filtered ahead. The filtered groups
are again written as \texttt{flowtools} files using the same \texttt{zlib}
compression level. The ratio of the number of output filtered groups to the
number of the input group is plotted against time. The results are shown in
Fig. \ref{fig:benchmarks-groupfilter}.

@@

It can be seen that the timings of \ac{NFQL} are far apart from that of SiLK.
It is due to the drift already created by the grouper at the $1.0$ ratio in
the previous stage. As a result, the group filter comes into play only after
$300$ seconds, whereas SiLK's group filtering already starts just below $150$
seconds. Even if we normalize the graph, it can be observed that the \ac{NFQL}
group filter has a higher slope. This is because it is only executed once the
grouper returns, and therefore has to reiterate the groups to make a filtering
decision.

@@

It can be seen that the merger is the most performance critical stage of the
\ac{NFQL} pipeline thus far. It is due to the fact that the merger is working
on twice the number of flow records than any other previous stage. In
addition, each branch is writing the results of the filter, grouper and group
filter stage to \texttt{flowtools} files. As a result, the amount of disk I/O
involved is twice as much as well. Even though each branch is delegated to a
separate core using affinity masks, most of time is taken in writing these
results to the file. These results although look less promising, they are way
better than the original \ac{NFQL} merger implementation. The optimized merger
implementation takes advantage of sorted nature of filtered groups and
therefore can signifantly reduce the number of merger matches. It also writes
a merged group record to file only once despite the number of times it has
matched. Without these optimizations, running such queries on the merger would
keep the CPU churning for days without results.

The last set of queries attempt to stress the ungrouper stage. They reuse the
entire merger queries as is, but enable the ungrouper as well. This means,
that the ungrouper now attempts to unfold the merged groups returned by the
merger to write their member flow records to \texttt{flowtools} files.
However, since the merger receives each flow record as its own filtered group,
each merged group has only one member. As a result, the ungrouper ends up
rewriting the merged groups as \texttt{flowtools} files using the same
\texttt{zlib} compression level. The ratio of the number of result flow
records to twice the number of the flow records in the input trace is plotted
against time. The evaluation results are shown in Fig.
\ref{fig:benchmarks-ungrouper}.

@@

It can be seen that the evaluation behavior of the ungrouper is similar to
that shown by the merger. Since, the ungrouper also has to write the entire
merged groupset to file, as is previously done by merger, the execution engine
takes twice the amount of time. It is important to note, that SiLK does not
have such an equivalent ungrouping tool is therefore not considered in
this final evaluation.

### 2012-09-02 sections/implementation.tex
* implementation   * flow query intermediate format     * custom c library to read flow records     * dedicated functions for each kind of operation   * execution workflow     * faster filter (with graph)     * grouper implementation       * possible grouping approaches       * faster grouping in generic case (nested bsearch)       * faster grouping in specific case (no bsearch needed) (with graph)       * cooked netflow v5 records for group aggregations     * group filter implementation     * merger implementation       * merger iterator utility       * skipping iterator permutations on sorte

\subsection{Splitter}
\subsection{Filter}
\subsection{Grouper}
\subsubsection{Group Aggregations}
\subsection{Group Filter}
\subsection{Merger}
\subsection{Ungrouper}

>>>>

The \ac{NFQL} architecture primarily consists of a front-end parser backed up
by an execution engine as shown in Fig. \ref{fig:architecture}. The execution
engine is the brain of \ac{NFQL} where the complete pipeline is processed. It
receives the flow query at runtime using a JSON intermediate format. The
execution engine is designed to receive this input from either a machine
running the front-end parser or a webservice that pushes the JSON from the
cloud. The execution engine is written in C, however the front-end parser can
be written in any desired language.

\begin{figure}[h!]
  \begin{center}
    \includegraphics* [width=1.0\linewidth]{nfql-architecture}
    \caption{NFQL Architecture}
    \label{fig:architecture}
  \end{center}
\end{figure}

\subsection{Flow Query Intermediate Format}
\input{sections/implementation/flow-query-intermediate-format}
\label{subsec:intermediate-format}

\subsection{Execution Workflow}
\input{sections/implementation/execution-workflow}
\label{subsec:execution-workflow}

\subsection{Performance Optimizations}
\input{sections/implementation/performance-optimizations}
\label{subsec:performance-optimizations}

\subsection{Adaptable Compression Levels}
\input{sections/implementation/adaptable-compression-levels}
\label{subsec:adaptable-compression-levels}

### 2012-09-02 sections/implementation/adaptable-compression-levels.tex
* implementation   * nfql architecture (with figure)   * flow query intermediate format     * citation to json-c     * flow query and branch structs (listing)     * efficient rule processing using dedicated comparators (listing)   * splitter   * filter     * nfql vs nfql (optimized) (figure)   * grouper     * nfql (generic) vs nfql (specific) (figure)     * group aggregations merged in this sub section   * group filter, merger, ungrouper   * performance evaluations

The flow-records echoed to the standard output can also be written to a
Netflow $v5$ flow-tools file. The \texttt{--dirpath} switch allows one to
provide a directory path where the results can be stored.  Each stream is
stored as its own file with an ID to disambiguate it. Results from each stage
of the pipeline can also be written to separate files with the increase in the
verbosity level. In fact, \texttt{--dirpath} and \texttt{--verbose} work well
together to adjust the level at which the writes are to be made.

@@

The engine uses the \texttt{zlib} \cite{rfc1950} software library to compress
the results written to the flow-tools files. \texttt{zlib} supports $9$
compression levels with $9$ being the highest. The \ac{NFQL} engine supports
\texttt{--zlevel} switch to allow the user to supply its desired choice of the
of the compression level. A default level of $5$ is used for writes if the
switch is not supplied during runtime.  Fig. \ref{fig:engine-zlevel} shows
the time taken to write a sample of records passing the filter stage for each
\texttt{z-level}. It goes to show that each level adds its own performance
overhead and must be used with discretion.

>>>>

The flow-records echoed to the standard output can also be written to a
Netflow $v5$ \texttt{flow-tools} file. The \texttt{--dirpath} switch allows
one to provide a directory path where the results can be stored.  Each stream
is stored as its own file with an ID to disambiguate it. Results from each
stage of the pipeline can also be written to separate files with the increase
in the verbosity level. In fact, \texttt{--dirpath} and \texttt{--verbose}
work well together to adjust the level at which the writes are to be made.

@@

The engine uses the \texttt{zlib} \cite{rfc1950} software library to compress
the results written to the \texttt{flow-tools} files. \texttt{zlib} supports
$9$ compression levels with $9$ being the highest. The \ac{NFQL} engine
supports \texttt{--zlevel} switch to allow the user to supply its desired
choice of the of the compression level. A default level of $5$ is used for
writes if the switch is not supplied during runtime.  Fig.
\ref{fig:engine-zlevel} shows the time taken to write a sample of records
passing the filter stage for each \texttt{z-level}. It goes to show that each
level adds its own performance overhead and must be used with discretion.

### 2012-09-02 sections/implementation/execution-workflow.tex
* implementation   * nfql architecture (with figure)   * flow query intermediate format     * citation to json-c     * flow query and branch structs (listing)     * efficient rule processing using dedicated comparators (listing)   * splitter   * filter     * nfql vs nfql (optimized) (figure)   * grouper     * nfql (generic) vs nfql (specific) (figure)     * group aggregations merged in this sub section   * group filter, merger, ungrouper   * performance evaluations

\subsubsection{Splitter} A custom C library was written to directly read/write
data in the \texttt{flow-tools} format. The library sequentially reads the
complete flow-records into memory to support random access required for
relative filtering. Each flow-record is stored in a \texttt{char} array and
the offsets to each field are stored in separate \texttt{structs}.  The array
of such records are indexed allowing fast retrieval in $O(1)$ time. This
allowed the flow-records to be easily referenced by an identifier, thereby
giving away the need to every time copy all the flow-records when moving ahead
in the processing pipeline.

In the default method, there is a comparison function defined for every
possible field length $(33)$ and comparison operations $(19)$. These functions
are generated using a Python script. The rule definitions are now able to make
calls using a function name derived from the combination of field length,
delta type and operation.  This subverts the need to define complex branching
statements and reduces complexity.

\subsubsection{Filter}
\begin{figure}[h!]
  \begin{center}
    \includegraphics* [width=1.0\linewidth]{filter-fv1-fv2}
    \caption{Filter Stage: F$(v1)$ vs F$(v2)$}
    \label{fig:fv1-fv2-filter}
  \end{center}
\end{figure}

The flow query language suggests to read all the flow records of a supplied
trace into memory before starting the processing pipeline.  Since, the filter
stage uses the supplied set of absolute rules to make a decision on whether or
not to keep a flow record; it has to pass through the whole in-memory
recordset \emph{again} to fill in the filter results. This technique involves
multiple linear runs on the trace and therefore slows down when the ratio of
number of filtered records to the total number of flow-records is high. The
\ac{NFQL} implementation merges the filter stage with in-memory read of the
trace. This means, a decision on whether or not to make room for a record in
memory and eventually hold a pointer for it in filter results is done upfront
as soon as the record is read from the trace. In addition, if a request to
write the filter stage results to a flow-tools file has been made, the writes
are also made as soon the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. It is important to note that the filtered records are
saved in common location from where they are referenced by each branch. This
helps keep the memory costs at a minimum when multiple branches are involved.
We used the publically available Trace 7 from the SimpleWeb \cite{simpleweb}
to compare the performance of the new filter in F$(v2)$ with that from F$(v1)$
as shown in Fig. \ref{fig:fv1-fv2-filter}. The trace has around $20M$
flow-records.

@@

In order to be able to make comparisons on field offsets, a simple approach is
to linearly walk through each filtered record against the filtered recordset
leading to a complexity of $O(n^2)$, where $n$ is the number of filtered
records. A smarter approach is to put the copy in a hash table and then try to
map each pointer while walking down the filtered recordset once, leading to a
complexity of $O(n)$. The hash table approach, although will work on this
specific example, will fail badly on other relative comparisons. A better
approach, as implemented in F$(v2)$ is to sort the filtered recordset on all
the requested grouping rules. This helps the execution engine perform a
nested \texttt{bsearch} to reduce the linear pass to a fairly small filtered
recordset. However, the number of elements upon which the search has to be
performed needs to be known at each level of the binary lookup. In order to
avoid a linear run to count the number of elements, the parent level
\texttt{bsearch} invocation returns a boundary with the first and last element
to enable such a calculation. This helps the grouper perform faster search
lookups to find records that must group together in $O(n*lg(k))$ time with a
preprocessing step taking $O(n*lg(n)) + O(n)$ in the average case, where $n$
is the number of filtered records and $k$ is the number of unique filtered
records.

\begin{figure}[h!]
  \begin{center}
    \includegraphics*[width=1.0\linewidth]{grouper-fv2-bsearch-fv2}
    \caption{Grouper Stage: F$(v2)$ (Generic) vs F$(v2)$ (Specific)}
    \label{fig:fv1-fv2-grouper}
  \end{center}
\end{figure}

The grouping approach has further been optimized when the filtered records are
grouped for equality. In such a scenario, the need to search for unique
records and a  subsequent binary search goes away.  The groups can now be
formed in $O(n)$ time with a more involved preprocessing step taking
$O(p*n*lg(n)$ where is $n$ is the number of filtered records, and $p$ is the
number of grouping rules. The performance evaluation of the grouper handling
this special case against its behavior when handling generic cases is shown in
Fig. \ref{fig:fv1-fv2-grouper}. Profiling the performance of the grouper on
queries that produce higher ratios reveal that a significant amount of time is
taken in \texttt{bsearch}. The special case of equality comparisons eliminate
these calls and further reduce the time in the long tail as shown in Fig.
\ref{fig:fv1-fv2-grouper}.

\subsubsection{Group Aggregations} Group records are a conglomeration of
several flow records with some common characteristics defined by the flow
query. Some of the non-common characteristics can also be aggregated into a
single value using group aggregations. It is useful if a single group record
can be again mapped into a NetFlow $v5$ record template, so that it can be
written to file as a representative of all its members.

\subsubsection{Group Filter} The groupfilter is used to filter the groups
produced by the grouper based on some absolute rules defined in a \ac{DNF}
expression. The \texttt{struct} term holds information about the flow record
offset, the value being compared to and the operator which maps to a unique
\texttt{enum} value.  This enum value is used to map the operation to a
specific group-filter function. The group-filter functions are auto-generated
using a python script.

@@

The merger is used to relate groups from different branches according to a
merging criterean. The implementation is not trivial since the number of
branches that need to be spawned is read from the query and is not known until
compile time. As a result, an iterator that can provide all possible
permutations of $m-$tuple (where $m$ is the number of branches) group record
$ID$s was needed.  The result of the iterator can then be used to make a
match. The merger stage, begins by initializing this iterator passing it the
number of branches, and information about each branch. Then, it loops over to
get a new $m-$tuple of group record $ID$s on each iteration until the iterator
returns \texttt{false}.

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(n^m)$ where $n$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records however can be reduced by sorting the group records on the field
offsets used for a match. The \ac{NFQL} implementation optimizes the merger to
skip over iterator permutations when a state of a current field offset value
may not allow any further match beyond the index in the current branch.  For
such an optimization to work, the filtered group records must be sorted in the
order of field offsets specified in the merger clause. Specifying the filtered
group records in any other order may lead to undefined behavior. This means,
that if the same field offsets were used in the grouper stage, the terms in
the group clause can be rearranged by the query designer to align with the
order of terms in the merger clause.

>>>>

A custom C library has been written to directly read/write data in the
\texttt{flow-tools} format. The library sequentially reads the complete
flow-records into memory to support random access required for relative
filtering. Each flow-record is stored in a \texttt{char} array and the offsets
to each field are stored in separate \texttt{structs} as shown in listing
\ref{lst:c-lib-struct}. The array of such records are indexed allowing fast
retrieval in $O(1)$ time.

\lstset{caption=Trace Data Struct,
				tabsize=2, language=C, numbers=left,stepnumber=1,
        basicstyle=\tiny\ttfamily, numberstyle=\ttfamily\color{gray},
        keywordstyle=\color{blue}, xleftmargin=15pt,
        rulesepcolor=\color{black}, label=lst:c-lib-struct,
        aboveskip=10pt, captionpos=b}
\begin{lstlisting}
struct ft_data {
  int                                         fd;
  struct ftio                                 io;
  struct fts3rec_offsets                      offsets;
  struct ftver                                version;
  u_int64_t                                   xfield;
  int                                         rec_size;
  char**                                      recordset;
  size_t                                      num_records;
};
\end{lstlisting}

A sample JSON query \ac{DNF} term definition is shown in listing
\ref{lst:filter-struct}. In order to be able to make comparisons on the field
offsets of such a term, the comparator needs to know the type of the
comparison and the length of the field offset. This information is parsed by
execution engine once the query is read and is therefore not available at
compile time.  In order to subvert the need to define complex branching
statements, a dedicated comparator is defined for every possible field length
and comparison operation. These comparators are generated using a Python
script. This allows the term definitions to make runtime calls using a
function name derived from the combination of operation type and field length.

\lstset{caption=Filter Term Struct,
				tabsize=2, language=C, numbers=left,stepnumber=1,
        basicstyle=\tiny\ttfamily, numberstyle=\ttfamily\color{gray},
        keywordstyle=\color{blue}, xleftmargin=15pt,
        rulesepcolor=\color{black}, label=lst:filter-struct,
        aboveskip=10pt, captionpos=b}
\begin{lstlisting}
struct filter_term {
  size_t                                      field_offset;
  uint64_t                                    value;
  uint64_t                                    delta;
  struct filter_op*                           op;
  bool (*func)(
               const char* const              record,
               size_t                         field_offset,
               uint64_t                       value,
               uint64_t                       delta
              );
};
\end{lstlisting}

\subsubsection{Splitter} \ac{NFQL} uses identifiers to reference a flow record
in the \texttt{char} array. This gives away the need to every time copy all
the flow-records when moving ahead in the processing pipeline. As a result,
there is no dedicated splitter stage in the execution engine. Each branch
references the flow records from a common memory location. This helps keep the
memory costs at a minimum when multiple branches are involved.

\subsubsection{Filter}
\begin{figure}[h!]
  \begin{center}
    \includegraphics* [width=1.0\linewidth]{filter-fv1-fv2}
    \caption{Filter Stage: NFQL vs NFQL (optimized)}
    \label{fig:fv1-fv2-filter}
  \end{center}
\end{figure}

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to fill in the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \ac{NFQL} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a flow-tools file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. We used the publically available Trace 7 from the
SimpleWeb \cite{simpleweb} to compare the performance of the optimized
\ac{NFQL} implementation against the one defined by the flow query language as
shown in Fig. \ref{fig:fv1-fv2-filter}. The filter stage implementation with
these optimizations runs 10 times faster and is more pronounced on higher
ratios.

@@

In order to be able to make relative comparisons on field offsets, a simple
approach is to linearly walk through each filtered record against the filtered
recordset leading to a complexity of $O(n^2)$, where $n$ is the number of
filtered records. A smarter approach is to put the copy in a hash table and
then try to map each pointer while walking down the filtered recordset once,
leading to a complexity of $O(n)$. The hash table approach however, will only
work on equality comparisons. A better approach, as implemented in \ac{NFQL}
is to sort the filtered recordset on all the requested grouping terms of a
clause. This helps the execution engine perform a nested binary search to
reduce the linear pass to a fairly small filtered recordset. However, the
number of elements upon which the search has to be performed needs to be known
at each level of the binary lookup. In order to avoid a linear run to count
the number of elements, the parent level binary search invocation returns a
boundary with the first and last element to enable such a calculation. This
helps the grouper perform faster search lookups to find records that must
group together in $O(n*lg(k))$ time with a preprocessing step taking
$O(p*n*lg(n))$ in the average case, where $n$ is the number of filtered
records, $p$ is the number of grouping terms in a clause and $k$ is the number
of unique filtered records.

\begin{figure}[h!]
  \begin{center}
    \includegraphics*[width=1.0\linewidth]{grouper-fv2-bsearch-fv2}
    \caption{Grouper Stage: NFQL (Generic) vs NFQL (Specific)}
    \label{fig:fv1-fv2-grouper}
  \end{center}
\end{figure}

The grouping approach has further been optimized when the filtered records are
grouped for equality. In such a scenario, the need to search for unique
records and a  subsequent binary search goes away.  The groups can now be
formed in $O(n)$ time with the same preprocessing step taking $O(p*n*lg(n)$.
The performance evaluation of the grouper handling this special case against
its behavior when handling generic cases is shown in Fig.
\ref{fig:fv1-fv2-grouper}. Profiling the performance of the grouper on queries
that produce higher ratios reveal that a significant amount of time is taken
in binary search which is eliminated in the special case of equality
comparisons.

The resultant group records are a conglomeration of multiple flow records with
some common characteristics. Some of the non-common characteristics can also
be aggregated into a single value using group aggregations as defined in the
query. Such an aggregated group record is again mapped to a NetFlow $v5$
record template. This allows the aggregated group records to be written to
file as a representative of all its members.

\subsubsection{Group Filter} The groupfilter stage is used to filter the
groups produced by the grouper based on some absolute rules defined in a
\ac{DNF} expression. The \texttt{struct} term holds information about the flow
record offset, the value being compared to and the operator which maps to a
unique \texttt{enum} value.  This enum value is used to map the operation to a
specific group-filter function. The group-filter functions are auto-generated
using a Python script.

@@

The merger is used to relate groups from different branches according to a
merging criterean. The implementation is not trivial since the number of
branches that need to be spawned is read from the query and is not known until
runtime. As a result, an iterator that can provide all possible permutations
of $m-$tuple (where $m$ is the number of branches) group record $ID$s was
needed.  The result of the iterator can then be used to make a match. The
merger stage, begins by initializing this iterator passing it the number of
branches, and information about each branch. Then, it loops over to get a new
$m-$tuple of group record $ID$s on each iteration until the iterator returns
\texttt{false}.

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records however can be reduced by sorting the group records on the field
offsets used for a match. The \ac{NFQL} implementation optimizes the merger to
skip over iterator permutations when a state of a current field offset value
may not allow any further match beyond the index in the current branch.  For
such an optimization to work, the filtered group records must be sorted in the
order of field offsets specified in the merger clause. Specifying the filtered
group records in any other order may lead to undefined behavior. This means,
that if the same field offsets were used in the grouper stage, the terms in
the group clause can be rearranged by the query designer to align with the
order of terms in the merger clause.
